# Wildfire Trends in the US: Section 1
## By Noa Mills

**This section of code downloads the Monitoring Trends in Burn Severity (MTBS) fire data, the National Oceanic and Atmospheric Administration (NOAA) weather station data, and identifies the nearest station to each fire that records each of the following key weather variables for analysis: Temperature Maximum (TMAX), Temperature Minimum (TMIN), and Precipitation (PRCP).**

First, install and load the necessary packages and set a time variable to track how long this script takes to run.

```{r echo=T, results='hide', error=FALSE, message=FALSE, warning=FALSE}
ptm <- proc.time()

if(!require(rio)){install.packages("rio")}
library(rio)
if(!require(dplyr)){install.packages("dplyr")}
library(dplyr)
if(!require(tidyr)){install.packages("tidyr")}
library(tidyr)
if(!require(rgdal)){install.packages("rgdal")}
library(rgdal)
if(!require(base)){install.packages("base")}
library(base)
if(!require(profvis)){install.packages("profvis")}
library(profvis)
if(!require(rnoaa)){install.packages("rnoaa")}
library(rnoaa)
if(!require(lubridate)){install.packages("lubridate")}
library(lubridate)
if(!require(sf)){install.packages("sf")}
library(sf)
if(!require(spData)){install.packages("spData")}
library(spData)
if(!require(tidyverse)){install.packages("tidyverse")}
library(tidyverse)
if(!require(data.table)){install.packages("data.table")}
library(data.table)
if(!require(ggplot2)){install.packages("ggplot2")}
library(ggplot2)
if(!require(foreach)){install.packages("foreach")}
library(foreach)
if(!require(doParallel)){install.packages("doParallel")}
library(doParallel)
if(!require(doMC)){install.packages("doMC")}
library(doMC)
```

Create directories to store data, and download the MTBS weather data. The file mtbs.csv contains data on fires in the contiguous US, Alaska, Hawaii, and Puerto Rico, including fires larger than 500 acres to the east of the Mississippi, and fires 1000 acres or larger to the west of the Mississippi. The data also includes some smaller fires. Data attributes include fire ID, date of ignition, longitude, latitude, and area burned.
```{r}
if(!dir.exists("Data")){
  dir.create("Data")
}
if(!dir.exists("Data/mtbs")){
  dir.create("Data/mtbs")
}
if(!file.exists('Data/mtbs/mtbs_data.csv')){
  loc <- "https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/MTBS_Fire/data/composite_data/fod_pt_shapefile/mtbs_fod_pts_data.zip"
  dest <- paste0('Data/mtbs/mtbs_data.zip')
  download.file(loc, dest)
  unzip(dest, exdir="Data/mtbs")
  unlink(dest)
  convert('Data/mtbs/mtbs_FODpoints_DD.dbf', 'Data/mtbs/mtbs_data.csv')
  #remove unnecessary files from zip
  file.remove('Data/mtbs/mtbs_FODpoints_DD.cpg')
  file.remove('Data/mtbs/mtbs_FODpoints_DD_metadata.xml')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.prj')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.sbn')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.sbx')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.shp')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.shp.xml')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.shx')
  file.remove('Data/mtbs/mtbs_FODpoints_DD.dbf')
}
```

Read the mtbs.csv data into a dataframe called 'firedata', convert the ignition date from a factor to a date, and extract year, month, and date values.
```{r}
firedata <- read.csv('Data/mtbs/mtbs_data.csv')

firedata$IG_DATE <- as.Date(firedata$Ig_Date, format = '%Y-%m-%e')
firedata$IG_YEAR <- as.numeric(format(firedata$IG_DATE, '%Y'))
firedata$IG_MONTH <- as.numeric(format(firedata$IG_DATE, '%m'))
firedata$IG_DAY <- as.numeric(format(firedata$IG_DATE, '%d'))
```

Download two data files from NOAA: inventory.txt and stations.txt. Code to read the .txt files into a dataframe is modified from [this documentation](http://spatialreasoning.com/wp/20170307_1244_r-reading-filtering-weather-data-from-the-global-historical-climatology-network-ghcn). The stations.txt file contains one row per station, and lists the ID, long and lat, state, and elevation for each station. The inventory.txt file contains a variable number of rows per station. Each row specifies a station ID, long and lat, a specific weather variable that was recorded at that station (such as TMAX, TMIN or PRCP), and the beginning and end dates of observation for that variable at that station.

*Potential next step: Incorporate elevation data into dataframe, and incorporate in models.*

```{r}
if(!dir.exists("Data/noaa")){
  dir.create("Data/noaa")
}
if(!dir.exists("Data/noaa/noaadata")){
  dir.create("Data/noaa/noaadata")
}
#download noaa weather stations file
if(!file.exists('Data/noaa/stations.txt')){
  download.file("https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt", 'Data/noaa/stations.txt')
}

#download noaa weather inventory file
if(!file.exists('Data/noaa/inventory.txt')){
  download.file("https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-inventory.txt", 'Data/noaa/inventory.txt')
}

#create stations dataframe
typedcols <- c( "A11", "F9", "F10", "F7", "X1","A2",
                "X1","A30", "X1", "A3", "X1", "A3", "X1", "A5" )
stations <- read.fortran("Data/noaa/stations.txt",
                         typedcols, 
                         comment.char="")
hdrs <- c("ID", "LAT", "LON", "ELEV", "ST", "NAME","GSN", "HCN", "WMOID")
names(stations) <- hdrs

#create inventory dataframe
invcols <- c( "A11", "X1", "F8", "X1", "F9", "X1","A4",
              "X1","I4", "X1", "I4" )
inv <- read.fortran("Data/noaa/inventory.txt",
                    invcols,
                    comment.char="")
invhdrs <- c("ID", "LAT", "LON", "ELEM" , "FIRST", "LAST")
names(inv) <- invhdrs
```

In this analysis, we only include three weather variables: TMAX, TMIN and PRCP. Here, we filter the inventory dataframe to only include rows that record the start and stop dates of one of these three weather variables.

```{r}
inv$ELEM <- as.character(inv$ELEM)
inv <- inv[which(inv$ELEM == "TMAX" | inv$ELEM == "TMIN" | inv$ELEM == "PRCP"),]
```

Next, we filter the stations dataframe to only include stations which are in the US or Puerto Rico, and save this data into the dataframe stationsUS

```{r}
statelist <- c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL", "GA", "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA", "MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE", "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA", "PR", "RI", "SC", "SD", "TN", "TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY")
stationsUS<-stations[which(stations$ST %in% statelist),]
```

Remove from the inventory dataframe any stations that are not in the stationsUS dataframe.

```{r}
inv$inStnUS <- inv$ID %in% stationsUS$ID
inv <- inv[which(inv$inStnUS),]
```

Remove from the inventory dataframe any stations that have never recorded weather variables from one of the three key variables used in this portfolio (TMAX, TMIN, PRCP).

*Potential improvement: make this code run faster by either using an apply function, or parallelizing the loop.*

```{r}
for (stn in 1:nrow(stationsUS)){
  if(!"TMAX" %in% inv[which(inv$ID == stationsUS$ID[stn]),]$ELEM |
     !"TMIN" %in% inv[which(inv$ID == stationsUS$ID[stn]),]$ELEM |
     !"PRCP" %in% inv[which(inv$ID == stationsUS$ID[stn]),]$ELEM){
    inv <- inv[which(inv$ID != stationsUS$ID[stn]),]
  }
}
```

Remove from the stations dataframe any stations that are not in the inventory dataframe, and thus do not record any of the key variables.

```{r}
stationsUS$inInv <- stationsUS$ID %in% inv$ID
stationsUS <- stationsUS[which(stationsUS$inInv),]
```

Add attributes to the stationsUS dataframe to indicate the start and stop dates of variable observations.

*Potential improvement: make this code, and the code in the next block, run faster by either using an apply function or parallelizing the loop.*

```{r}
stationsUS$startTMAX <- 0
stationsUS$startTMIN <- 0
stationsUS$startPRCP <- 0
stationsUS$endTMAX <- 0
stationsUS$endTMIN <- 0
stationsUS$endPRCP <- 0

for (stn in 1:nrow(stationsUS)){
  stationsUS$startTMAX[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "TMAX"),]$FIRST[1]
  stationsUS$endTMAX[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "TMAX"),]$LAST[1]
  stationsUS$startTMIN[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "TMIN"),]$FIRST[1]
  stationsUS$endTMIN[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "TMIN"),]$LAST[1]
  stationsUS$startPRCP[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "PRCP"),]$FIRST[1]
  stationsUS$endPRCP[stn] <- inv[which(inv$ID == stationsUS[stn,]$ID & inv$ELEM == "PRCP"),]$LAST[1]
}
```

Then, add an attribute to the stationsUS dataframe indicating the first and last dates of observation.
```{r}
stationsUS$LAST <- 0
for(stn in 1:nrow(stationsUS)){
  stationsUS[stn,]$LAST <- inv[which(inv$ID == stationsUS[stn,]$ID),][1,]$LAST
}

stationsUS$FIRST <- 0
for(stn in 1:nrow(stationsUS)){
  stationsUS[stn,]$FIRST <- inv[which(inv$ID == stationsUS[stn,]$ID),][1,]$FIRST
}
```

Remove from the stationsUS dataframe all stations whose attribute 'LAST' was before 1984, indicating that the given station was decomissioned before the timeframe of the MTBS wildfire data.

Save the stationsUS dataframe as a csv.

```{r}
stationsUS$decom <- FALSE
for(i in 1:nrow(inv)){
  if(inv[i,]$LAST < 1984){
    stationsUS[which(stationsUS$ID == inv[i,]$ID),]$decom = TRUE
  }
}
stationsUS <- stationsUS[which(stationsUS$decom == FALSE),]
write.csv(stationsUS, "Data/noaa/stationsUS.csv")
```

Split the firedata by fire type. For the main analysis here, we only observe fires that are Wildfires, Wildland Use Fires, and Prescribed Burns. All other or unknown fires are stored into firedataOther.csv.

```{r}
otherFiredata <- firedata[which(!firedata$Incid_Type %in% c("Prescribed Fire", "Wildfire", "Wildland Fire Use")),]
#remove from firedata all fires that are not wildfires
firedata <- firedata[which(firedata$Incid_Type %in% c("Wildfire", "Prescribed Fire", "Wildland Fire Use")),]
#write.csv(firedata, "Data/mtbs/firedata1.csv")
write.csv(otherFiredata, "Data/mtbs/otherFiredata.csv")
```

Now, we find the nearest weather stations to each fire and incorporate it in the firedata dataframe. Since not all stations observe all three of the key weather variables over the same time periods, we identify the nearest stations for TMAX, TMIN and PRCP separately. First, we create dummy variables to represent the long and lat of the nearest station, and the distance of the station to the given fire.

```{r}
#add closest station fields to mtbs firedata
#fields initiated as a dummy station located outside the US
firedata$closestStnDist_TMAX <- 10000
firedata$closestStnID_TMAX <- "dummy station"
firedata$closestStnLong_TMAX <- 90
firedata$closestStnLat_TMAX <- -30
firedata$closestStnDist_TMIN <- 10000
firedata$closestStnID_TMIN <- "dummy station"
firedata$closestStnLong_TMIN <- 90
firedata$closestStnLat_TMIN <- -30
firedata$closestStnDist_PRCP <- 10000
firedata$closestStnID_PRCP <- "dummy station"
firedata$closestStnLong_PRCP <- 90
firedata$closestStnLat_PRCP <- -30
```

This function approximates the distance in kilometers between to long/lat points, and was borrowed from [this source](https://www.google.com/url?q=https://conservationecology.wordpress.com/2013/06/30/distance-between-two-points-in-r/&sa=D&ust=1590514403133000&usg=AFQjCNFQnMg9-YSmAMVK-nLN2eHnJSEQTQ)

```{r}
earth.dist <- function (long1, lat1, long2, lat2){
  rad <- pi/180
  a1 <- lat1 * rad
  a2 <- long1 * rad
  b1 <- lat2 * rad
  b2 <- long2 * rad
  dlon <- b2 - a2
  dlat <- b1 - a1
  a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  R <- 6378.145
  d <- R * c
  d <- round(d, 5)
  return(d)
}
```

Now, we identify the nearest station for each fire, and for each variable. The initial 'dummy' long/lat values for closest stations are far away from the US. For each fire, we iterate through the entire list of stations. For each station, if the distance from that station to the fire is less than the distance of the currently recorded closest station, then we update the closest station. This code is parallelized with the foreach package, and would run faster on a compute cluster. The firedata dataframe is saved as Data/firedata1.csv.
  

```{r}
registerDoParallel()
firedataOut <- foreach (row=1:nrow(firedata), .combine = rbind) %dopar% {
  newRow <- firedata[row,]
  for (stn in 1:nrow(stationsUS)){
    #find closest station with TMAX observations during time period
    prevDist <- newRow$closestStnDist_TMAX[1]
    newDist <- earth.dist(newRow$BurnBndLon[1], newRow$BurnBndLat[1], stationsUS$LON[stn], stationsUS$LAT[stn])
    if (newRow$IG_YEAR[1] <= stationsUS$endTMAX[stn] & newRow$IG_YEAR[1] >= stationsUS$startTMAX[stn] &
        newDist < prevDist){
      newRow$closestStnDist_TMAX[1] <- newDist
      newRow$closestStnID_TMAX[1] <- toString(stationsUS$ID[stn])
      newRow$closestStnLong_TMAX[1] <-  stationsUS$LON[stn]
      newRow$closestStnLat_TMAX[1] <-  stationsUS$LAT[stn]
    }
    
    #TMIN
    prevDist <- newRow$closestStnDist_TMIN[1]
    newDist <- earth.dist(newRow$BurnBndLon[1], newRow$BurnBndLat[1], stationsUS$LON[stn], stationsUS$LAT[stn])
    if (newRow$IG_YEAR[1] <= stationsUS$endTMIN[stn] & newRow$IG_YEAR[1] >= stationsUS$startTMIN[stn] &
        newDist < prevDist){
      newRow$closestStnDist_TMIN[1] <- newDist
      newRow$closestStnID_TMIN[1] <- toString(stationsUS$ID[stn])
      newRow$closestStnLong_TMIN[1] <-  stationsUS$LON[stn]
      newRow$closestStnLat_TMIN[1] <-  stationsUS$LAT[stn]
    }
    
    #PRCP
    prevDist <- newRow$closestStnDist_PRCP[1]
    newDist <- earth.dist(newRow$BurnBndLon[1], newRow$BurnBndLat[1], stationsUS$LON[stn], stationsUS$LAT[stn])
    if (newRow$IG_YEAR[1] <= stationsUS$endPRCP[stn] & newRow$IG_YEAR[1] >= stationsUS$startPRCP[stn] &
        newDist < prevDist){
      newRow$closestStnDist_PRCP[1] <- newDist
      newRow$closestStnID_PRCP[1] <- toString(stationsUS$ID[stn])
      newRow$closestStnLong_PRCP[1] <-  stationsUS$LON[stn]
      newRow$closestStnLat_PRCP[1] <-  stationsUS$LAT[stn]
    }
  }
  newRow
}


firedata <- firedataOut

write.csv(firedata, "Data/firedata1.csv")
totalTime <- proc.time() - ptm
print(paste0("Total time: ", round(totalTime[3]/60, digits=5), " minutes"))
```