# Wildfire Trends in the US: Section 3
## By Noa Mills

**In this script, we identify the geographic region and state of each fire, and calculate averages of weather variables for each fire across various timeframes. We also create dummy (boolean) variables so we can use the categorical variables present in the dataframe in our linear regression model.**

First, as before, we load the needed packages and the firedata dataframe.

```{r, message=FALSE, warning=FALSE}
library(rio)
library(dplyr)
library(tidyr)
library(rgdal)
library(base)
library(profvis)
library(rnoaa)
library(lubridate)
library(sf)
library(spData)
library(tidyverse)
library(data.table)
library(ggplot2)


files <- list.files(path="Data/", pattern="firedata2c_[0123456789]+.csv", recursive=FALSE)
iteration <- 1
for(file in files){
  fileval <- gsub("firedata2c_", "", file)
  fileval <- gsub(".csv", "", fileval)
  fileval <- as.integer(fileval)
  if(fileval > iteration){
    iteration <- fileval
  }
}
firedata <- read.csv(paste0("Data/firedata2b_", iteration, ".csv"), stringsAsFactors = FALSE)
```

Next, we remove any fires from the dataframe that fail to meet our threshold requirements for the numbers of flagged or missing data observations. In scripts 2a, 2b and 2c we iterated through nearest stations to find the nearest station with enough good data. However, if we exceeded a distance of 250 km away from the fire, we stopped. This code block ensures that those fires are disregarded as they aren't associated with relevant weather data. When run with MTBS data up until 2018, only one fire was removed from the firedata dataset in this code block.

```{r}
minFlag = 10
minNA = 15
firedata <- firedata %>% filter(flagTMAX < minFlag & flagTMIN < minFlag & flagPRCP < minFlag & naTMAX < minNA & naTMIN < minNA & naPRCP < minNA)

```

MTBS identifies the assessment type of each fire as Initial or Extended. Initial assessments were conducted soon after the fire in ecosystems such as grasslands that experience quick regeneration, whereas extended assessments were conducted in ecosystems such as forests and woodlands where the effects of fire persist beyond the next growing season. Thus, assessment type is a rough proxy for land type.

If only a single image was used to perform the assessment, then the assessment type is appended with the abbreviation "SS" which stands for single source. Fires without this abbreviation were assessed with both before and after images when constructing a digital fire perimeter. When building our model to predict wildfire size, we care much more about the assessment type than the number of images used to conduct the assessment, so here we remove the "SS" abbreviations to merge single source and multi source assessment types into the same category.

```{r}
firedata[which(firedata$Asmnt_Type == "Initial (SS)"),]$Asmnt_Type <- "Initial"
firedata[which(firedata$Asmnt_Type == "Extended (SS)"),]$Asmnt_Type <- "Extended"
```

Now we download shapefiles from the National Weather Service that delineates state boundaries.

```{r}
if(!dir.exists("Data/nws")){
  dir.create("Data/nws")
}

if(!file.exists('Data/nws/s_11au16.shp')){
  download.file("https://www.weather.gov/source/gis/Shapefiles/County/s_11au16.zip", 'Data/nws/states_and_territories.zip')
  unzip('Data/nws/states_and_territories.zip', exdir = "Data/nws")
  unlink('Data/nws/states_and_territories.zip')
}

states_and_territories <- st_read("Data/nws/s_11au16.shp")
states_and_territories %>%
  filter(!NAME %in% c("American Samoa","Guam", "Northern Marianas", "Virgin Islands"))
```

We use a function provided by [this stackoverflow post](https://stackoverflow.com/questions/8751497/latitude-longitude-coordinates-to-state-code-in-r) to identify state membership based on longitude and latitude.

```{r}
## pointsDF: A data.frame whose first column contains longitudes and
##           whose second column contains latitudes.
##
## states:   An sf MULTIPOLYGON object with 50 states plus DC. //editted to use nws multipolygon
##
## name_col: Name of a column in `states` that supplies the states'
##           names.
lonlat_to_state <- function(pointsDF,
                            states = states_and_territories,
                            name_col = "NAME") {
  ## Convert points data.frame to an sf POINTS object
  pts <- st_as_sf(pointsDF, coords = 1:2, crs = 4326)
  
  ## Transform spatial data to some planar coordinate system
  ## (e.g. Web Mercator) as required for geometric operations
  states <- st_transform(states, crs = 3857)
  pts <- st_transform(pts, crs = 3857)
  
  ## Find names of state (if any) intersected by each point
  state_names <- states[[name_col]]
  ii <- as.integer(st_intersects(pts, states))
  state_names[ii]
}
#end of code block borrowed from stack to convert long/lat to state
```

We use this function to identify the state of each fire.
```{r}
lonlatvals<-select(firedata, BurnBndLon, BurnBndLat)
firedata$state <- lonlat_to_state(lonlatvals)
```

Using the MTBS data up until 2018, after applying this function to identify statehood, there are still 36 fires whose state is not identified, and where the function returns NA. This is often because the epicenter of the fire is just outside of the US boundary, or is in a body of water. So, I created maps of each of these fires and manually identify the state that the fire is in. Then, I created a vector of state names that corresponds to the fires, and updated the state names accordingly. This code only needs to be run (and the vector of state names manually updated) each time new data is released from MTBS, so it is commented out in the code file, and is not evaluated here.

```{r, eval=FALSE}
if(!dir.exists("Output/stateFinder")){
  dir.create("Output/stateFinder")
}
firedataStateNA <- firedata[which(is.na(state)),]
for(i in 1:nrow(firedataStateNA)){
  firedata_row <- firedataStateNA[i,]
  pointLon <- firedata_row$BurnBndLon[1]
  pointLat <- firedata_row$BurnBndLat[1]
  plot <- ggplot(data=states_and_territories)+
    geom_sf()+
    geom_point(data=firedata_row, mapping = aes(x=BurnBndLon, y=BurnBndLat), color="red",alpha=0.8)+
    coord_sf(xlim=c(pointLon - 8,pointLon + 8), ylim=c(pointLat - 6, pointLat + 6))+
    #scale_size_continuous(range=c(2,9))+
    theme_void()
  ggsave(paste0("Output/stateFinder/map", i, ".png"), plot, width=15)
}

firedataStateNA$state <- c("Washington", "Minnesota", "California", "New Mexico", "Florida", "New Mexico", "Florida", "Florida",
                           "Minnesota", "Montana", "California", "Alaska", "Alaska", "Florida", "Florida", "Alaska",
                           "California", "Florida", "Minnesota", "North Carolina", "Florida", "Florida", "Florida", "Florida", 
                           "Florida", "Florida", "New Mexico", "Arizona", "California", "Minnesota", "North Carolina", "Alaska",
                           "Florida", "Arizona", "Texas", "New Mexico")
detach(firedata)

#make sure none of the states were spelled wrong
print(sort(unique(union(firedata$state, firedataStateNA$state))))

for(i in 1:nrow(firedataStateNA)){
  firedata[which(firedata$Event_ID == firedataStateNA$Event_ID[i]),] <- firedataStateNA[i,]
  print(paste0("Replaced: ", firedataStateNA$Event_ID[i], " state with ", firedataStateNA$state[i]))
}
```

Now, we identify the geographic region of each fire based on state membership. These regions are defined by the National Weather Service.

```{r}
firedata$region <- "region"
firedata[which(firedata$state %in% c("California", "Nevada")),]$region <- "West"
firedata[which(firedata$state %in% c("Utah", "Colorado", "Arizona", "New Mexico")),]$region <- "Southwest"
firedata[which(firedata$state %in% c("Washington", "Oregon", "Idaho")),]$region <- "Northwest"
firedata[which(firedata$state %in% c("Montana", "North Dakota", "South Dakota", "Wyoming", "Nebraska")),]$region <- "Northern Rockies and Plains"
firedata[which(firedata$state %in% c("Kansas", "Oklahoma", "Texas", "Arkansas", "Louisiana", "Mississippi")),]$region <- "South"
firedata[which(firedata$state %in% c("Minnesota", "Wisconsin", "Michigan", "Iowa")),]$region <- "Upper Midwest"
firedata[which(firedata$state %in% c("Missouri", "Illinois", "Indiana", "Ohio", "West Virginia", "Kentucky", "Tennessee")),]$region <- "Ohio Valley"
firedata[which(firedata$state %in% c("Virginia", "North Carolina", "South Carolina", "Georgia", "Alabama", "Florida")),]$region <- "Southeast"
firedata[which(firedata$state %in% c("Maryland", "Delaware", "New Jersey", "Pennsylvania", "Connecticut", "Rhode Island", "New York", "Vermont", "Massachusetts", "New Hampshire", "Maine", "DC")),]$region <- "Northeast"
firedata[which(firedata$state == "Alaska"),]$region <- "Alaska"
firedata[which(firedata$state == "Hawaii"),]$region <- "Hawaii"
firedata[which(firedata$state == "Puerto Rico"),]$region <- "Puerto Rico"
```

Next we create dummy (boolean) variables for fire type, assessment type, region, month, and year group (set of 5 years). This allows us to include this categorical data in our linear regression model.

```{r}
#Dummy variables for fire type
firedata$isRx <- 0
firedata$isWf <- 0
firedata$isWFU <- 0
firedata[which(firedata$Incid_Type == "Wildfire"),]$isWf <- 1
firedata[which(firedata$Incid_Type == "Prescribed Fire"),]$isRx <- 1
firedata[which(firedata$Incid_Type == "Wildland Fire Use"),]$isWFU <- 1

#Dummy variables for assessment type
firedata$isInitial <- 0
firedata$isExtended <- 0
firedata[which(firedata$Asmnt_Type == "Initial"),]$isInitial <- 1
firedata[which(firedata$Asmnt_Type == "Extended"),]$isExtended <- 1

#Dummy variables for region
firedata$isSW <- 0
firedata$isSE <- 0
firedata$isS <- 0
firedata$isW <- 0
firedata$isNRP <- 0
firedata$isOV <- 0
firedata$isNW <- 0
firedata$isNE <- 0
firedata$isUM <- 0
firedata$isAL <- 0
firedata$isPR <- 0
firedata$isHI <- 0
firedata[which(firedata$region == "Southwest"),]$isSW <- 1
firedata[which(firedata$region == "Southeast"),]$isSE <- 1
firedata[which(firedata$region == "South"),]$isS <- 1
firedata[which(firedata$region == "West"),]$isW <- 1
firedata[which(firedata$region == "Northern Rockies and Plains"),]$isNRP <- 1
firedata[which(firedata$region == "Ohio Valley"),]$isOV <- 1
firedata[which(firedata$region == "Northwest"),]$isNW <- 1
firedata[which(firedata$region == "Northeast"),]$isNE <- 1
firedata[which(firedata$region == "Upper Midwest"),]$isUM <- 1
firedata[which(firedata$region == "Alaska"),]$isAL <- 1
firedata[which(firedata$region == "Puerto Rico"),]$isPR <- 1
firedata[which(firedata$region == "Hawaii"),]$isHI <- 1

#Dummy variables for month
firedata$isJan <- 0
firedata$isFeb <- 0
firedata$isMar <- 0
firedata$isApr <- 0
firedata$isMay <- 0
firedata$isJun <- 0
firedata$isJul <- 0
firedata$isAug <- 0
firedata$isSep <- 0
firedata$isOct <- 0
firedata$isNov <- 0
firedata$isDec <- 0
firedata[which(firedata$IG_MONTH == 1),]$isJan <- 1
firedata[which(firedata$IG_MONTH == 2),]$isFeb <- 1
firedata[which(firedata$IG_MONTH == 3),]$isMar <- 1
firedata[which(firedata$IG_MONTH == 4),]$isApr <- 1
firedata[which(firedata$IG_MONTH == 5),]$isMay <- 1
firedata[which(firedata$IG_MONTH == 6),]$isJun <- 1
firedata[which(firedata$IG_MONTH == 7),]$isJul <- 1
firedata[which(firedata$IG_MONTH == 8),]$isAug <- 1
firedata[which(firedata$IG_MONTH == 9),]$isSep <- 1
firedata[which(firedata$IG_MONTH == 10),]$isOct <- 1
firedata[which(firedata$IG_MONTH == 11),]$isNov <- 1
firedata[which(firedata$IG_MONTH == 12),]$isDec <- 1
firedata$isFireSeason <- 0
firedata[which(firedata$IG_MONTH %in% c(7,8,9,10)),]$isFireSeason <- 1

#Dummy variables for year
#Separate into groups by 5 years
#84-88, 89-93, 94-98, 99-03, 04-08, 09-13, 14-18
firedata$year_group <- ""
firedata[which(firedata$IG_YEAR >=1984 & firedata$IG_YEAR <= 1988),]$year_group <- "1984-1988"
firedata[which(firedata$IG_YEAR >=1989 & firedata$IG_YEAR <= 1993),]$year_group <- "1989-1993"
firedata[which(firedata$IG_YEAR >=1994 & firedata$IG_YEAR <= 1998),]$year_group <- "1994-1998"
firedata[which(firedata$IG_YEAR >=1998 & firedata$IG_YEAR <= 2003),]$year_group <- "1998-2003"
firedata[which(firedata$IG_YEAR >=2004 & firedata$IG_YEAR <= 2008),]$year_group <- "2004-2008"
firedata[which(firedata$IG_YEAR >=2009 & firedata$IG_YEAR <= 2013),]$year_group <- "2009-2013"
firedata[which(firedata$IG_YEAR >=2014 & firedata$IG_YEAR <= 2018),]$year_group <- "2014-2018"
```

Lastly, we calculate averages of the weather variables.
ELMavg represents the average across the entire timespan (3 weeks before until 5 weeks after ignition)
ELMavg.i represents the weekly average i weeks before ignition
ELMavgi represents the weekly average i weeks after ignition

```{r}

firedata$TMAXavg <- rowMeans(firedata %>% select(starts_with("TMAX")), na.rm=TRUE)
firedata$prevTMAXavg <- rowMeans(firedata[,c("TMAX.3", "TMAX.2", "TMAX.1", "TMAX0", 
                                              "TMAX1", "TMAX2", "TMAX3")], na.rm=TRUE)
firedata$TMAXavg.3 <- rowMeans(firedata[,c("TMAX.21", "TMAX.20", "TMAX.19", "TMAX.18", 
                                           "TMAX.17", "TMAX.16", "TMAX.15")], na.rm=TRUE)
firedata$TMAXavg.2 <- rowMeans(firedata[,c("TMAX.14", "TMAX.13", "TMAX.12", "TMAX.11", 
                                           "TMAX.10", "TMAX.9", "TMAX.8")], na.rm=TRUE)
firedata$TMAXavg.1 <- rowMeans(firedata[,c("TMAX.7", "TMAX.6", "TMAX.5", "TMAX.4", 
                                           "TMAX.3", "TMAX.2", "TMAX.1")], na.rm=TRUE)
firedata$TMAXavg0 <- rowMeans(firedata[,c("TMAX1", "TMAX2", "TMAX3", "TMAX4", 
                                           "TMAX5", "TMAX6", "TMAX7")], na.rm=TRUE)
firedata$TMAXavg1 <- rowMeans(firedata[,c("TMAX8", "TMAX9", "TMAX10", "TMAX11", 
                                          "TMAX12", "TMAX13", "TMAX14")], na.rm=TRUE)
firedata$TMAXavg2 <- rowMeans(firedata[,c("TMAX15", "TMAX16", "TMAX17", "TMAX18", 
                                          "TMAX19", "TMAX20", "TMAX21")], na.rm=TRUE)
firedata$TMAXavg3 <- rowMeans(firedata[,c("TMAX22", "TMAX23", "TMAX24", "TMAX25", 
                                          "TMAX26", "TMAX27", "TMAX28")], na.rm=TRUE)
firedata$TMAXavg4 <- rowMeans(firedata[,c("TMAX29", "TMAX30", "TMAX31", "TMAX32", 
                                          "TMAX33", "TMAX34", "TMAX35")], na.rm=TRUE)

firedata$TMINavg <- rowMeans(firedata %>% select(starts_with("TMIN")), na.rm=TRUE)
firedata$prevTMINavg <- rowMeans(firedata[,c("TMIN.3", "TMIN.2", "TMIN.1", "TMIN0", 
                                              "TMIN1", "TMIN2", "TMIN3")], na.rm=TRUE)
firedata$TMINavg.3 <- rowMeans(firedata[,c("TMIN.21", "TMIN.20", "TMIN.19", "TMIN.18", 
                                           "TMIN.17", "TMIN.16", "TMIN.15")], na.rm=TRUE)
firedata$TMINavg.2 <- rowMeans(firedata[,c("TMIN.14", "TMIN.13", "TMIN.12", "TMIN.11", 
                                           "TMIN.10", "TMIN.9", "TMIN.8")], na.rm=TRUE)
firedata$TMINavg.1 <- rowMeans(firedata[,c("TMIN.7", "TMIN.6", "TMIN.5", "TMIN.4", 
                                           "TMIN.3", "TMIN.2", "TMIN.1")], na.rm=TRUE)
firedata$TMINavg0 <- rowMeans(firedata[,c("TMIN1", "TMIN2", "TMIN3", "TMIN4", 
                                          "TMIN5", "TMIN6", "TMIN7")], na.rm=TRUE)
firedata$TMINavg1 <- rowMeans(firedata[,c("TMIN8", "TMIN9", "TMIN10", "TMIN11", 
                                          "TMIN12", "TMIN13", "TMIN14")], na.rm=TRUE)
firedata$TMINavg2 <- rowMeans(firedata[,c("TMIN15", "TMIN16", "TMIN17", "TMIN18", 
                                          "TMIN19", "TMIN20", "TMIN21")], na.rm=TRUE)
firedata$TMINavg3 <- rowMeans(firedata[,c("TMIN22", "TMIN23", "TMIN24", "TMIN25", 
                                          "TMIN26", "TMIN27", "TMIN28")], na.rm=TRUE)
firedata$TMINavg4 <- rowMeans(firedata[,c("TMIN29", "TMIN30", "TMIN31", "TMIN32", 
                                          "TMIN33", "TMIN34", "TMIN35")], na.rm=TRUE)
firedata$PRCPavg <- rowMeans(firedata %>% select(starts_with("PRCP")), na.rm=TRUE)
firedata$prevPRCPavg <- rowMeans(firedata[,c("PRCP.3", "PRCP.2", "PRCP.1", "PRCP0", 
                                              "PRCP1", "PRCP2", "PRCP3")], na.rm=TRUE)
firedata$PRCPavg.3 <- rowMeans(firedata[,c("PRCP.21", "PRCP.20", "PRCP.19", "PRCP.18", 
                                           "PRCP.17", "PRCP.16", "PRCP.15")], na.rm=TRUE)
firedata$PRCPavg.2 <- rowMeans(firedata[,c("PRCP.14", "PRCP.13", "PRCP.12", "PRCP.11", 
                                           "PRCP.10", "PRCP.9", "PRCP.8")], na.rm=TRUE)
firedata$PRCPavg.1 <- rowMeans(firedata[,c("PRCP.7", "PRCP.6", "PRCP.5", "PRCP.4", 
                                           "PRCP.3", "PRCP.2", "PRCP.1")], na.rm=TRUE)
firedata$PRCPavg0 <- rowMeans(firedata[,c("PRCP1", "PRCP2", "PRCP3", "PRCP4", 
                                          "PRCP5", "PRCP6", "PRCP7")], na.rm=TRUE)
firedata$PRCPavg1 <- rowMeans(firedata[,c("PRCP8", "PRCP9", "PRCP10", "PRCP11", 
                                          "PRCP12", "PRCP13", "PRCP14")], na.rm=TRUE)
firedata$PRCPavg2 <- rowMeans(firedata[,c("PRCP15", "PRCP16", "PRCP17", "PRCP18", 
                                          "PRCP19", "PRCP20", "PRCP21")], na.rm=TRUE)
firedata$PRCPavg3 <- rowMeans(firedata[,c("PRCP22", "PRCP23", "PRCP24", "PRCP25", 
                                          "PRCP26", "PRCP27", "PRCP28")], na.rm=TRUE)
firedata$PRCPavg4 <- rowMeans(firedata[,c("PRCP29", "PRCP30", "PRCP31", "PRCP32", 
                                          "PRCP33", "PRCP34", "PRCP35")], na.rm=TRUE)

write.csv(firedata, "Data/firedata3.csv")
```